---
title: "Biostat 203B Homework 5"
subtitle: Due Mar 20 @ 11:59PM
author: "Sakshi Oza, 606542442"
format:
  html:
    theme: cosmo
    embed-resources: true
    number-sections: false
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
---

## Predicting ICU duration

Using the ICU cohort `mimiciv_icu_cohort.rds` you built in Homework 4, develop at least three machine learning approaches (logistic regression with enet regularization, random forest, boosting, SVM, MLP, etc) plus a model stacking approach for predicting whether a patient's ICU stay will be longer than 2 days. You should use the `los_long` variable as the outcome. You algorithms can use patient demographic information (gender, age at ICU `intime`, marital status, race), ICU admission information (first care unit), the last lab measurements before the ICU stay, and first vital measurements during ICU stay as features. You are welcome to use any feature engineering techniques you think are appropriate; but make sure to not use features that are not available at an ICU stay's `intime`. For instance, `last_careunit` cannot be used in your algorithms. 
```{r}
library(bigrquery)
library(dbplyr)
library(DBI)
library(gt)
library(gtsummary)
library(tidyverse)
```

```{r}
# path to the service account token
satoken <- "./biostat-203b-2025-winter-4e58ec6e5579.json"
# BigQuery authentication using service account
bq_auth(path = satoken)
```

```{r}
# connect to the BigQuery database `biostat-203b-2025-mimiciv_3_1`
con_bq <- dbConnect(
  bigrquery::bigquery(),
  project = "biostat-203b-2025-winter",
  dataset = "mimiciv_3_1",
  billing = "biostat-203b-2025-winter"
)
con_bq
```


```{r}
library(tidyverse)
library(tidymodels)

data <- readRDS("./mimic_icu_cohort.rds")

```


1. Data preprocessing and feature engineering.
```{r}
# Load necessary libraries
library(tidyverse)
library(tidymodels)
library(rsample)
library(ggplot2)
library(gtsummary)
library(dplyr)
library(vip)
library(doParallel)
library(xgboost)
library(ranger)
library(stacks)
library(future)


# Preprocess the data
mimic_icu_cohort <- mimic_icu_cohort %>%
  mutate(los_long = ifelse(los > 2, 1, 0)) %>%
  select(-los, -last_careunit) %>%  # Remove los and last_careunit
  drop_na()  # Remove rows with missing values

# Sort the data by subject_id, hadm_id, and stay_id
mimic_icu_cohort <- mimic_icu_cohort %>%
  arrange(subject_id, hadm_id, stay_id)



```


2. Partition data into 50% training set and 50% test set. Stratify partitioning according to `los_long`. For grading purpose, sort the data by `subject_id`, `hadm_id`, and `stay_id` and use the seed `203` for the initial data split. Below is the sample code.
```{r}
# Set seed for reproducibility
set.seed(203)

# Stratified partitioning
split <- initial_split(mimic_icu_cohort, prop = 0.5, strata = los_long)
train_data <- training(split)
test_data <- testing(split)
```

3. Train and tune the models using the training set.
```{r}
# Define the recipe for preprocessing
recipe <- recipe(los_long ~ ., data = train_data) %>%
  step_impute_mean(all_numeric_predictors()) %>%
  step_impute_mode(all_nominal_predictors()) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors()) %>%
  step_normalize(all_numeric_predictors())

# Define models
logistic_model <- logistic_reg(penalty = tune(), mixture = tune()) %>%
  set_engine("glmnet")

rf_model <- rand_forest(mtry = tune(), trees = 1000) %>%
  set_engine("ranger") %>%
  set_mode("classification")

xgb_model <- boost_tree(tree_depth = tune(), learn_rate = tune()) %>%
  set_engine("xgboost") %>%
  set_mode("classification")

# Create a workflow set
models <- workflow_set(
  preproc = list(recipe),
  models = list(logistic = logistic_model, rf = rf_model, xgb = xgb_model)
)

# Train models using cross-validation
set.seed(203)
folds <- vfold_cv(train_data, v = 5)

results <- models %>%
  workflow_map("tune_grid", resamples = folds, grid = 10)

# Select the best model based on ROC AUC
best_results <- results %>%
  extract_workflow_set_result("recipe_logistic") %>%
  select_best(metric = "roc_auc")

final_model <- finalize_workflow(
  extract_workflow(results, "recipe_logistic"),
  best_results
) %>%
  fit(train_data)
```



4. Compare model classification performance on the test set. Report both the area under ROC curve and accuracy for each machine learning algorithm and the model stacking. Interpret the results. What are the most important features in predicting long ICU stays? How do the models compare in terms of performance and interpretability?

